---
title: "The Economic Cost of Weather Events"
subtitle: "A historical analysis of the economic impacts of severe weather events in the United States from 1950 to 2011."
author: "Derek Q. Jackson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

# Synopsis

-----

# Data Processing
This section outlines the process of acquiring and preparing the NOAA Storm Database for analysis. The raw data, provided as a compressed *.csv.bz2* file, is read directly from the source website using the *fread* function from the *data.table* package, which efficiently handles large compressed files. The dataset contains records of major storms and weather events in the United States from 1950 to November 2011, including details on event types, locations, fatalities, injuries, and property damage.

## Package Installation & Loading
First, it's important to ensure that all necessary R packages for running this analysis & report are installed and loaded. If you are running this analysis for the first time, these packages will be installed automatically.

```{r package_installation, cache=TRUE}
required_packages <- c("data.table", "lubridate", "ggplot2", "cowplot", "reshape2")

# Check if packages are installed, install if not, then load them
for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}
```

## Data Acquisition
The NOAA Storm Database is accessed directly from the URL without saving it locally. The fread function reads the compressed *.csv.bz2* file into a *data.table* object for efficient processing. To optimize performance, caching is enabled for this time-consuming operation.

```{r download_data, cache=TRUE}
options(knitr.cache = TRUE)

data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
storm_data <- fread(data_url, stringsAsFactors = FALSE)
```

## Data Cleaning & Preparation
The raw NOAA Storm Database requires cleaning to ensure data quality, consistency, and usability for addressing the project's key questions: identifying weather events most harmful to population health (via fatalities and injuries) and those with the greatest economic consequences (via property and crop damage). Without proper cleaning, inconsistencies such as varying case in event types, missing values, or improperly formatted damage estimates could lead to inaccurate aggregations, biased results, or errors in the analysis. 

For instance, the database spans decades with evolving recording practices, leading to potential data entry variations and incompleteness in earlier years. The cleaning steps are designed to mitigate these issues while preserving as much valid data as possible, ensuring reproducibility, and aligning with the National Weather Service's documentation on variable definitions. The following subsections detail each step with its rationale and corresponding code.

**1. Inspecting the Dataset**

Summarizing key columns *(EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG)* helps understand the data's structure, ranges, and potential issues like outliers or data types. We also explicitly check for missing values in these critical columns to quantify data gaps and inform handling strategies, ensuring we make informed decisions about data retention or removal.

```{r}
# Inspect the dataset
summary(storm_data[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG)])

# Check for missing values in critical columns
missing_summary <- storm_data[, .(
  EVTYPE_missing = sum(is.na(EVTYPE)),
  FATALITIES_missing = sum(is.na(FATALITIES)),
  INJURIES_missing = sum(is.na(INJURIES)),
  PROPDMG_missing = sum(is.na(PROPDMG)),
  CROPDMG_missing = sum(is.na(CROPDMG))
)]
print(missing_summary)
```

The output of *missing_summary* indicates no missing values *(NA)* in the critical columns *(EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG)*. This suggests the dataset is complete for these variables, which simplifies our cleaning process. However, we proceed with checks for other potential issues, such as empty strings or invalid entries, in subsequent steps.

**2. Handling Missing or Invalid Values**

Although the inspection shows no *NA* values in the critical columns, we include a step to handle potential edge cases, such as empty strings in *EVTYPE* or zero values in damage columns that might reflect unreported data. For *EVTYPE*, we ensure no empty strings *(e.g., "")* exist, as this column is central to grouping events for analysis. Rows with empty *EVTYPE* strings would be meaningless for aggregating health or economic impacts and are removed. For *PROPDMG* and *CROPDMG*, values of zero are retained, as the documentation suggests unreported damage implies no significant loss, but we verify that no negative values exist, which would be invalid for damage estimates.

```{r}
# Remove rows with empty or whitespace-only EVTYPE
storm_data <- storm_data[EVTYPE != "" & !is.na(trimws(EVTYPE))]

# Verify no negative values in damage columns
storm_data[PROPDMG < 0, PROPDMG := 0]
storm_data[CROPDMG < 0, CROPDMG := 0]
```

**3. Standardizing Event Types & Removing Duplicate Entries**

The *EVTYPE* column may contain inconsistencies due to manual entry over many years, such as variations in casing (e.g., "Tornado" vs. "TORNADO"), extra spaces, or minor misspellings. Converting to uppercase and trimming whitespace reduces these duplicates, allowing accurate aggregation by event type.

```{r}
# Standardize EVTYPE
storm_data[, EVTYPE := toupper(trimws(EVTYPE))]
```

The EVTYPE column also contains near-duplicate entries due to abbreviations or slight variations in naming (e.g., "TSTM WIND" and "THUNDERSTORM WIND" both refer to winds produced by thunderstorms). Combining these duplicates is essential to avoid underestimating the impact of similar events in aggregations for population health or economic consequences. Based on the official event types from the National Weather Service Storm Data Documentation, we map common variations to standardized official names. Note that "FLOOD" and "FLASH FLOOD" are kept separate, as they are distinct official event types (general flooding vs. sudden flooding), unless specified otherwise.

```{r}
# Map common duplicates to official event types
storm_data[grepl("TSTM WIND|THUNDERSTORM WINDS|THUNDERSTORMWIND|THUNDERTORM WIND|THUDERSTORM WINDS|THUNDERSTROM WIND|THUNDERSTORM WIND.*|TSTMW|THUNDERSTORMW|THUNDERSTORM WINDS.*", EVTYPE), EVTYPE := "THUNDERSTORM WIND"]
storm_data[grepl("HIGH WIND.*|STRONG WIND.*|GUSTY WIND", EVTYPE), EVTYPE := "HIGH WIND"]
storm_data[grepl("EXCESSIVE HEAT|HEAT WAVE|HEATWAVE|EXTREME HEAT", EVTYPE), EVTYPE := "EXCESSIVE HEAT"]
storm_data[grepl("WINTER STORM|WINTER WEATHER|MIXED PRECIP|WINTRY MIX|BLIZZARD|ICE STORM", EVTYPE), EVTYPE := "WINTER STORM"]
storm_data[grepl("FLASH FLOOD.*|FLASH FLOODING", EVTYPE), EVTYPE := "FLASH FLOOD"]
storm_data[grepl("FLOOD.*|URBAN FLOOD|COASTAL FLOOD", EVTYPE), EVTYPE := "FLOOD"]
```

**4. Processing Damage Variables**

Property and crop damage are recorded in two parts: a numeric base *(PROPDMG, CROPDMG)* and an exponent code *(PROPDMGEXP, CROPDMGEXP)* such as *"K" (thousands)*, *"M" (millions)*, or *"B" (billions)*, as per the Storm Data Documentation. Without conversion, these cannot be summed meaningfully for total economic consequences. 

We define a function to map exponents to numeric multipliers (defaulting to 1 for empty or invalid codes to avoid data loss) and compute *TOTAL_DMG* in consistent USD units. This step ensures comparable economic impacts across events.

```{r}
# Define function to convert damage exponents to numeric multipliers
convert_exp <- function(exp) {
  exp <- toupper(exp)
  ifelse(exp == "K", 1e3,
         ifelse(exp == "M", 1e6,
                ifelse(exp == "B", 1e9, 1)))
}

# Convert damage exponents and calculate total damage in USD
storm_data[, PROPDMGEXP := ifelse(is.na(PROPDMGEXP) | PROPDMGEXP == "", 1, convert_exp(PROPDMGEXP))]
storm_data[, CROPDMGEXP := ifelse(is.na(CROPDMGEXP) | CROPDMGEXP == "", 1, convert_exp(CROPDMGEXP))]
storm_data[, TOTAL_DMG := PROPDMG * PROPDMGEXP + CROPDMG * CROPDMGEXP]
```

**5. Parsing Dates**

The BGN_DATE column is in a string format *(e.g., "MM/DD/YYYY HH:MM:SS")*, which is not suitable for chronological analysis or filtering. Using *lubridate::mdy_hms*, we convert it to a POSIXct date-time object with UTC timezone to enable time-based insights, such as trends over years or seasonal patterns.

```{r}
# Parse BGN_DATE to date format
storm_data[, BGN_DATE := mdy_hms(BGN_DATE, tz = "UTC")]
```

### Verifying the Cleaned Data
After cleaning, we verify the transformed dataset by showing *head* of the key variables to ensure the transformations were applied correctly. This step confirms that event types are standardized, health and damage metrics are complete, and dates are properly formatted, setting the stage for accurate analysis of population health and economic impacts.

```{r}
# Verify cleaned data
head(storm_data[, .(EVTYPE, FATALITIES, INJURIES, TOTAL_DMG, BGN_DATE)])
```

-----

# Analysis
This section presents the analysis of the NOAA Storm Database to address two questions: 

**(1) Which types of weather events are most harmful to population health?** 
**(2) Which types have the greatest economic consequences?**

## Population Health Impacts
To determine the weather events most harmful to population health, we aggregate *FATALITIES* and *INJURIES* by __*EVTYPE*__ to capture both fatal and non-fatal impacts. We select the top 10 event types based on total injuries and visualize both fatalities and injuries within each bar to provide a comprehensive view of health risks. The bars are ordered by the total number of combined injuries and fatalities, with fatalities shown in orange and injuries in blue.

```{r}
# Aggregate fatalities and injuries by event type
health_impact <- storm_data[, .(
  Total_Fatalities = sum(FATALITIES),
  Total_Injuries = sum(INJURIES)
), by = EVTYPE]

# Calculate combined health impact
health_impact[, Total_Health_Impact := Total_Fatalities + Total_Injuries]

# Get top 10 event types by combined health impact
top_health <- health_impact[order(-Total_Health_Impact)][1:10]

# Reshape data for combined bar plot
health_long <- melt(top_health[, .(EVTYPE, Total_Fatalities, Total_Injuries)], 
                    id.vars = "EVTYPE", measure.vars = c("Total_Fatalities", "Total_Injuries"),
                    variable.name = "Impact_Type", value.name = "Count")
```

```{r, health_damage_plot, fig.cap="This bar plot displays the health impacts (Injuries and Fatalities) of the top 10 weather event types."}
# Create combined bar plot ordered by combined health impact from largest to smallest
p1 <- ggplot(health_long, aes(x = factor(EVTYPE, levels = rev(top_health$EVTYPE)), y = Count, fill = Impact_Type)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = c("Total_Fatalities" = "#D55E00", "Total_Injuries" = "#0072B2"),
                    labels = c("Fatalities", "Injuries")) +
  coord_flip() +
  labs(x = "Event Type", y = "Total Count", title = "Top 10 Weather Events - Health Impacts",
       fill = "Impact Type") +
  scale_y_continuous(labels = scales::comma, breaks = scales::pretty_breaks(n = 5)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    axis.text = element_text(color = "black"),
    axis.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(color = "black", linewidth = 1),
    plot.margin = unit(c(0.5, 1.5, 0.5, 0.5), "cm")  # Ensure no cutoff
  )

# Display the plot
print(p1)
```
### Figure 1: Top Weather Events Impacting Population Health
This bar plot displays the top 10 weather event types by total injuries in the NOAA Storm Database (1950–2011), with each bar showing both injuries (blue) and fatalities (orange) stacked together. The events are ordered by the total number, highlighting tornadoes and thunderstorm winds as leading causes of harm, while also revealing significant harm counts for floods and excessive heat.

## Economic Consequences

-----

# Results

-----