---
title: "Reproducible Research Project 1"
author: "Derek Q. Jackson"
output: html_document
---

---

## Introduction

This report presents an analysis of personal activity monitoring data collected from an anonymous individual during **October and November 2012.** The data set, comprising 5-minute interval measurements, includes variables such as the number of steps taken, the date, and the specific 5-minute interval identifier.

The data set can be downloaded from: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

The variables included in this data set are:

- **steps:** Number of steps taking in a 5-minute interval (missing values are coded as 
*NA*)

- **date:** The date on which the measurement was taken in YYYY-MM-DD format

- **interval:** Identifier for the 5-minute interval in which measurement was taken

This analysis aims to explore daily activity patterns, quantify total steps, investigate average daily activity trends, address missing data, and compare activity levels between weekdays and weekends.

The project emphasizes reproducible research practices. All data loading, processing, analysis, and visualization steps are conducted using R and documented within this R Markdown file. This ensures that the entire analysis can be fully replicated, promoting transparency and verifying the integrity of the findings presented herein.

---

#### Loading & Exploring the Data

Load the data set into a data table directly. Then view the head of the data.

```{r}
activityData <<- data.table::fread(file = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip")

head(activityData)
```

The head shows that there are many NAs in the steps. The date column is in IDate format.
The interval column is in integer format.

Let's look at a summary of the data.

```{r}
summary(activityData, na.rm = TRUE)
```

We see here that the the minimum number of steps taken in any 5 minute interval is 0.
The maximum number of steps taken in any 5 minute interval is 806. 

Interval is key here, because it is represented as integers. These integers actually correspond
to a time of day, i.e. 0 means 12 AM (0000 in 24-hour format) and 2355 means 1155 PM (2355 in 
24-hour format).

---

#### Pre-Processing

1. Convert interval to chron time.

2. Create a combined PosixCT field that contains a combination of the date and interval.

---

#### What is the Mean Total Number of Steps Taken Per Day?

1. Calculate the total number of steps taken per day.

2. Here is a histogram of the total number of steps taken each day.

3. Calculate the mean and median of the total number of steps taken per day.

---

#### What is the Average Daily Activity Pattern?

1. Here is a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

###### Which 5-minute interval, on average across all the days in the data set, contains the maximum number of steps?

---

#### Imputing Missing Values

Note that there are days/intervals where there are missing values (coded as *NA*).
The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate the total number of missing values in the data set

2. Fill in the missing values with the mean for that day.

3. Create a new data set that is equal to the original with the missing data filled in.

4. Here is a histogram of the total number of steps taken each day, with the imputed values.

5. Here is the mean and median number of steps taken per day.

###### Do these values differ from the estimates from the original summary?

###### What is the impact of imputing missing data on the estimates of the total daily number of steps?

---

#### Are there differences in Activty Patterns between Weekdays & Weekends?

1. Create a new factor variable in the data set with two levels ("weekday" and "weekend") indicating whether a given date is a weekday or weekend day.

2. Here is a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

---